Environmental warming increases the importance of high-turnover energy channels in stream food webs

### 2022-12-23 doc/ms/Junker_temp-energy-flux_submission_v2.Rmd
getting ready to transfer over to new repo and submit

---
title: "Environmental warming increases the importance of high-turnover energy channels in stream food webs"
author:
  - James R. Junker:
      email: jrjunker@mtu.edu
      institute: [msu, glrc]
      correspondence: true
  - name: Wyatt F. Cross
    institute: msu
  - name: James M. Hood
    institute: osu
  - name: Jonathan P. Benstead
    institute: bama
  - name: Alexander D. Huryn
    institute: bama
  - name: Daniel Nelson
    institute: usu
  - name: Jón S. Ólafsson
    institute: veidi
  - name: Gísli M. Gíslason
    institute: reyk
institute:
  - msu: Department of Ecology, Montana State University, Bozeman, MT 59717, USA
  - glrc: Current `address:` Great Lakes Research Center, Michigan Technological University, Houghton, MI 49931, USA
  - osu: Department of Evolution, Ecology, and Organismal Biology, The Ohio State University, Translational Data Analytics Institute, The Aquatic Ecology Laboratory, Columbus, OH 43212, USA
  - bama: Department of Biological Sciences, University of Alabama, Tuscaloosa, AL  35487, USA
  - usu: National Aquatic Monitoring Center, Department of Watershed Sciences, Utah State University, Logan, UT 84322, USA
  - veidi: Institute of Marine and Freshwater Fisheries, Hafnarfjördur, Iceland
  - reyk: University of Iceland, Institute of Life and Environmental Sciences, Reykjavík, Iceland
output:
   word_document:
      reference_docx: docx_template.docx
      toc: no
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--filter=pandoc-crossref.exe'
   pdf_document:
     keep_tex: true
   html_document: default
   pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--filter=pandoc-crossref.exe'
header-includes:
- \usepackage{lineno}
- \usepackage{amsmath}
- \numberwithin{equation}
- \usepackage{indentfirst}
- \linenumbers
indent: true
linestretch: 1
bibliography: refs.bib
link-citations: no
link-color: grey
csl: ecology.csl
---

@@

Key words: climate change; energy flux; food webs; species traits; warming

@@

@@

Warming temperatures are altering communities and trophic networks globally. While the overall influence of warming on food webs is often context-dependent, increasing temperatures are predicted to change communities in two fundamental ways: 1) by reducing average body size and 2) by increasing individual metabolic rates. These warming-induced changes have the potential to influence the distribution of food web fluxes, food web stability, and the relative importance of ecological and stochastic processes shaping community assembly. Here, we quantified patterns and the relative distribution of organic matter fluxes through stream food webs spanning a natural ~25$^\circ$C temperature gradient. We then related these patterns to species- and community-level trait distributions of mean body size and population biomass turnover (*P:B*) within and across ecosystems. We predicted that 1) communities in warmer streams would skew towards smaller body size and higher *P:B* and 2) organic matter fluxes within warmer communities would increasingly skew towards smaller, higher *P:B* populations. Across the temperature gradient, communities in warmer streams were on average composed of smaller-bodied and higher *P:B* populations. Additionally, material fluxes *within* warmer streams were increasingly skewed towards high-turnover populations, demonstrating that warming can restructure organic matter fluxes in both an absolute and relative sense. With warming, the relative distribution of organic matter fluxes also appeared to be increasingly 'non-random', suggesting the stronger selection for traits driving high turnover with increasing temperature. Our study suggests that a warming world will be 'smaller and faster', and that warming has the potential to shape pathways of energy and material flows in food webs in ways that may be predictable.

@@

<!--- warming and effects in food webs --->
<!--- Speeding up of food web: Species levels--->
Increasing global temperatures are influencing the provision and maintenance of ecosystem services by modifying the network of species interactions that underpin ecosystem functions [@brose2012; @woodward2010; @deruiter1995; @thompson2012]. The effects of warming permeate across levels of biological organization from its control on individual metabolic rates [@gillooly2001; @brown2004] and biological activity (e.g., attack rate, handling time, growth rates, etc.; @dell2014, to broad-scale shifts in community assembly and structure [@saito2021; @gibert2019a; @nelson2017a]. Across global climate gradients, such temperature-induced changes have the potential to mediate food web stability [@baiser2019] by altering the acquisition and allocation of resources [@zhang2017] and the magnitude and relative distribution of energy fluxes in food webs [@may1972; @mccann1998; @barnes2018].

Understanding how warming alters energy fluxes in food webs requires information about how temperature modifies connections between ecosystem structure and function. Such connections hinge upon how warming influences functional trait distributions and how functional traits translate to energy demands [@loreau2001; @norberg2001; @norberg2004]. Additionally, warming may modify both deterministic [e.g., environmental/niche filtering, @whittaker1962] and stochastic [e.g., neutral theory, @hubbell2001] community assembly processes [@saito2021], leading to many possible relationships between warming and species abundance and trait distributions. For example, although temperature appears to have minimal control on species richness across taxonomic groups [e.g., @bastazini2021], warming may have profound effects on the dominance structure (i.e., evenness) of ecological communities, often by reducing community evenness and favoring a few well-adapted species [@hillebrand2008]. This strong environmental filtering is likely to skew trait distributions in natural communities [@therriault1999]. However, the relative distribution of traits can be modified further by additional processes beyond environmental filtering that alter species relative abundance distributions [e.g., species interactions, @therriault1999;\; demographic stochasticity, @hubbell2001]. These additional processes can exaggerate or counter any skew in trait distributions imparted through environmental filtering and thereby modify the relationship between species' traits and the absolute and relative energy demands in food webs.

<!--- Connect traits to the 'speed' of ecosystem energy fluxes--->
Body size is a fundamental trait that is strongly influenced by temperature and has great potential to influence energy flux [@atkinson1994; @gardner2011; @daufresne2009]. Body size reductions arise from multiple mechanisms acting at different levels of organization, including the increased relative abundance of smaller species within warmer communities [@bergmann1848] or smaller individuals in warmer populations [@james1970], or reduced absolute body size of warmer individuals [@atkinson1994]. These temperature-size relationships are likely to influence the distribution of body sizes both across and within ecosystems, such that warmer communities should contain smaller species and individuals, on average. These relationships can have important implications for ecosystems patterns and processes that control energy and material fluxes in ecosystems because body size is a strong determinant of species life-history patterns [@peters1983; @zeuss2017; @altermatt2010; @nelson2020b], as well as developmental [@angilletta2004] and metabolic rates [@brown2004; @gillooly2001].

<!--- Write the main goals of the paragraph--->
Metabolic rate, in addition to being controlled by body size, is another key trait that is influenced by temperature and tied to variation in energy flux through food webs [@gillooly2001]. Warming can influence metabolic rates indirectly through reductions in body size, as well as directly through its effects on subcellular kinetics [@bideault2019; @osmond2017]. These processes have been shown to modify ecosystem patterns through changes in population abundance [@bernhardt2018], consumer-resource interactions [@bideault2019], and food web structure [@gibert2019a]. Importantly, the kinetic effects of temperature are tied to many biological processes [@dell2014], among them growth rate [@gillooly2001], developmental rate [@zuo2012; @nelson2020b], voltinism [@zeuss2017], and population turnover rate [@huryn2007; @brown2004]. Taken together, these changes suggest that warming should lead to a smaller, faster world in which ecosystem processes 'speed up' through the effects of smaller body size and higher turnover rates, with a potentially strong imprint on food web structure and energy fluxes [@gibert2019a].

We quantified patterns of organic matter flux in stream food webs across a natural temperature gradient (\~5--28$^\circ$C) in southwestern Iceland. Previous research in these streams has shown a strong positive effect of warming on primary production both among streams [@demars2011; @padfield2017] and seasonally within streams [@ogorman2012; @hood2018]. Invertebrates in these streams rely on autotrophs [@ogorman2012; @nelson2020], thus the dynamics of primary production functions as a strong control on energy flow through consumers [@junker2020]. We therefore predicted that annual organic matter fluxes to consumers would increase with stream temperature, mirroring patterns in resource availability and consumer energy demand. We also hypothesized that temperature would act as a principle environmental filter on community assembly and organic matter fluxes by favoring 'fast' life-history traits associated with small-bodied organisms. Specifically, we predicted that warming temperatures among streams would lead to a decrease in average body size and an increase in average biomass turnover (i.e., *P:B* ratio) of populations. We also predicted that *within* communities, organic matter fluxes would be dominated by small-bodied and high *P:B* taxa at higher temperatures and that these patterns would not arise by random change [i.e., a 'selection effect' *sensu* @huston1997]. Our results should help refine general predictions about how ongoing climate warming, and its influence on key traits, is likely to shape energy flux through animal food webs.

@@

We studied six streams within the Hengill geothermal field of southwestern Iceland (64$^\circ$ 03'N 021$^\circ$ 18'W) that varied in mean annual temperature from ~5 to 30$^\circ$C. Hengill is
characterized by indirect geothermal heating of groundwater [@arnason1969], leading to natural variability in stream temperatures (4.5--54.0 $^\circ$C), but similar solute
chemistries [@friberg2009]. These conditions create a "natural laboratory" for isolating the effects of temperature on ecosystem processes [@ogorman2014; @nelson2017a]. We selected streams to maximize the temperature range, while minimizing differences in the structural aspects of the primary producer community [@junker2020a]. In each stream, we measured temperature and water depth every 15 min from July 2010 through August 2012 (U20-001-01 water-level logger, Onset Computer Corp., Pocasset, MA, USA). Light availability in the watershed was measured every 15 min from atmospheric stations (HOBO pendant temperature/light UA-002-64, Onset Computer Corp., Pocasset, MA, USA).

@@

Daily secondary production of invertebrate taxa was calculated using the instantaneous growth rate method [IGR, @benke2017]. Growth rates were determined using taxon-appropriate
approaches described in Junker and others [-@junker2020]. Briefly, growth rates of common taxa (e.g., Chironomidae spp., *Radix balthica*, etc.) were determined using *in situ* chambers [@huryn1986].
Multiple individuals (*n* = 5--15) within small size categories (\~1-mm length range) were photographed next to a field micrometer, placed in the stream within pre-conditioned
chambers for 7--15 days, and removed and photographed. Individual lengths were measured from field pictures using image analysis software [@schindelin2012], and body
lengths were converted to mass (mg ash-free dry mass [AFDM]) using published length-mass regressions [@benke1999; @ogorman2012; @hannesdottir2013]. Growth rates (*g*, d^-1^) were
calculated by the changes in mean body size (*M*) over a given time interval (*t*) with the following equation:

@@

Organic matter fluxes through the community were calculated using the trophic basis of production method [TBP\; @benke1980]. Taxon-specific secondary production estimates were combined with diet proportions (see Supplemental Materials), diet-specific assimilation efficiencies, *AE~i~*, and net production efficiencies, *NPE* [e.g., @mccullough1975], to estimate consumption of organic matter. For each food
category, *i*, diet proportions were multiplied by the gross growth efficiency ($GGE_{i} = AE_{i} * NPE$) to calculate the relative production attributable to each food
category. The relative production from each food type was then multiplied by the interval-level production and finally divided by $GGE_{i}$ to estimate consumption of organic matter from each food category by each taxon [@benke1980]. Consumption was calculated for each taxon across sampling intervals (typically \~1 month). Total interval consumption was calculated by summing across all taxa, while annual consumption was calculated by summing across all taxa and intervals. Variability in consumption estimates was estimated through a Monte Carlo approach, wherein bootstrapped vectors of secondary production for each taxon (see *Secondary production* methods above) were resampled and consumption estimated with the TBP method using modeled diet proportions (see *Diet analysis* in Supplemental Materials), diet-specific assimilation efficiencies, and net production efficiency. Variability in *AE~i~* was incorporated by resampling values from beta distributions fit to median and 2.5% and 97.5% percentiles for each diet item: diatoms = 0.30 (95% percentile interval (PI): 0.24-0.36), filamentous and green algae = 0.30 (95% PI: 0.24-0.36), cyanobacteria = 0.10 (95% PI: 0.08-0.12), amorphous detritus = 0.10 (95% PI: 0.08-0.12), vascular and non-vascular plants (bryophytes) = 0.1 (95% PI: 0.08-0.12), and animal material = 0.7 (95% PI: 0.56-0.84)[@welch1968; @cross2007; @cross2011; @benke1980; @benke1997]. Variability in *NPE* was incorporated by resampling values from an assumed beta distribution with median *NPE* = 0.45 (95% PI = 0.4-0.5). Beta distributions were fit in R [@rcoreteam2022] using the 'get.beta.par()' function within the *rriskDistributions* package [@belgorodski2017].

@@

We predicted that warming would favor taxa with smaller body size and higher population turnover and therefore OM fluxes would be skewed towards small body size (*M*) and higher *P:B* across and within communities. To assess the potential effects of environmental filtering across communities, we used bootstrapped linear regressions to quantify the relationship between mean annual temperature ($^\circ$C) and 1) mean population body size ($\overline{M}$) and 2) biomass turnover ($\overline{P:B}$) of each stream community. Here, `r format(temperature_stats[['n_boot']], big.mark = ",", scientific = FALSE)` values of population *M* or *P:B* were resampled with replacement from population secondary production vectors (see *Secondary Production* above) for each taxon within each stream. The mean of all populations within a stream was calculated and a linear model was fit between *log~e~*-transformed $\overline{M}$ or $\overline{P:B}$ and mean annual temperature. Response variables were transformed to meet the assumption of normally distributed residual variation.

@@

To examine whether and how temperature influences the skew of OM flux among populations within each community, we quantified the probability of observing the skewed distributions in OM fluxes by random chance. The feasible range of skewness values within a community is inherently tied to the evenness of OM fluxes. This dependence on evenness can make it difficult to determine the importance of random versus ecological processes on the distribution of OM fluxes by comparison of raw skewness measures alone. We predicted that species $M$ and $P:B$ would be increasingly important traits structuring OM fluxes within a community and therefore warmer streams would exhibit highly skewed OM distributions that would be unlikely due to chance (i.e., 'non-random ordering'). In contrast, cooler streams would exhibit OM fluxes with skew values that are more likely due to random chance ('random ordering'), regardless of raw skewness values, suggesting other traits or processes govern the distribution of OM fluxes within their communities. To accomplish this, we first had to account for statistical constraints that restrict the range of possible outcomes [i.e., feasible set, @haegeman2008; @diaz2021a], given the number of species and the relative distribution of OM fluxes within a community. The number of unique orderings of species increases to computationally intractable numbers very quickly (e.g., $S!$ for 10 species $\approx 3.6 \times 10^{6}$ unique orderings). Therefore, we chose to permute a random subset of each stream community's feasible set by randomly ordering species and calculating skewness in the cumulative distribution of annual OM fluxes 100,000 times in each stream. This permuted set allowed us to calculate the probability of observing the empirical skewness value, $Sk_{flux}$, compared to a random ordering given the distribution of relative OM flux. We assessed the likelihood of non-random ordering as the distance from the central mass of the random skew distributions within each stream. Therefore, observed skew values near the tails of the random distribution where probabilities of observing $Sk_{flux}$ greater or equal ( Pr($Sk_{flux} \geq x$) ) are nearer to zero or one represent communities in which OM fluxes are likely organized non-randomly.

@@

We compared our skew estimates (i.e., $Sk_{flux}$) based on ordered data (i.e., ordered from small to large *M* or low to high *P:B* ratios; Figure 3B and Figure 4B) to skew estimates based on randomly sorted data to detect whether OM fluxes through smaller and higher *P:B* taxa could be attributed to random community assembly or likely based on these traits. The probability of observing a similar or more extreme skew of OM fluxes in relation to body size was variable among streams and ranged from `r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_mean")) %>% round(2)` (`r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_quant2.5")) %>% round(2)`--`r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_quant97.5")) %>% round(2)`; 95% PI) to `r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_mean")) %>% round(2)` (`r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_quant2.5")) %>% round(2)`--`r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_quant97.5")) %>% round(2)`); there was a very weak positive association between this probability and temperature (Figure 3C). In contrast, the probability of a similar or more extreme skew in relation to *P:B* ratios ranged from `r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_mean")) %>% round(2)` (`r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant2.5")) %>% round(2)`--`r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant97.5")) %>% round(2)`) to `r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_mean")) %>% round(2)` (`r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant2.5")) %>% round(2)`--`r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant97.5")) %>% round(2)`; Figure 4C) and in this case there was a clear trend towards a more structured OM flux distribution---i.e., favoring fluxes through high *P:B* taxa--at warmer temperatures. The likelihood of elevated fluxes among high *P:B* taxa became much more likely with warming, and the probability of random ordering decreased by `r round(pb_prob_temp_median, 3)*100`% (`r round(pb_prob_temp_quant2.5,3)*100`%--`r round(pb_prob_temp_quant97.5,3)*100`%) for every 1$^\circ$C increase in temperature (Figure 4C).

@@

@@

## Figure 1

@@

## Figure 2

>>>>

---
title: "Environmental warming increases the importance of high-turnover energy channels in stream food webs"
author:
  - James R. Junker:
      email: jrjunker@mtu.edu
      institute: [msu, glrc]
      correspondence: true
  - name: Wyatt F. Cross
    institute: msu
  - name: James M. Hood
    institute: osu
  - name: Jonathan P. Benstead
    institute: bama
  - name: Alexander D. Huryn
    institute: bama
  - name: Daniel Nelson
    institute: usu
  - name: Jón S. Ólafsson
    institute: veidi
  - name: Gísli M. Gíslason
    institute: reyk
institute:
  - msu: Department of Ecology, Montana State University, Bozeman, MT 59717, USA
  - glrc: Current `address:` Great Lakes Research Center, Michigan Technological University, Houghton, MI 49931, USA
  - osu: Department of Evolution, Ecology, and Organismal Biology, The Ohio State University, Translational Data Analytics Institute, The Aquatic Ecology Laboratory, Columbus, OH 43212, USA
  - bama: Department of Biological Sciences, University of Alabama, Tuscaloosa, AL  35487, USA
  - usu: National Aquatic Monitoring Center, Department of Watershed Sciences, Utah State University, Logan, UT 84322, USA
  - veidi: Institute of Marine and Freshwater Fisheries, Hafnarfjördur, Iceland
  - reyk: University of Iceland, Institute of Life and Environmental Sciences, Reykjavík, Iceland
output:
   word_document:
      reference_docx: working_docx_template.docx
      toc: no
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--filter=pandoc-crossref.exe'
   pdf_document:
     keep_tex: true
   html_document: default
   pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--filter=pandoc-crossref.exe'
header-includes:
- \usepackage{lineno}
- \usepackage{amsmath}
- \numberwithin{equation}
- \usepackage{indentfirst}
- \linenumbers
indent: true
linestretch: 1
bibliography: refs.bib
link-citations: no
link-color: grey
csl: ecology.csl
---

@@

@@

Key words: energy flux; climate change; food webs; species traits; warming

@@

Warming temperatures are altering communities and trophic networks globally. While the overall influence of warming on food webs is often context-dependent, increasing temperatures are predicted to change communities in two fundamental ways: 1) by reducing average body size and 2) by increasing individual metabolic rates. These warming-induced changes have the potential to influence the distribution of food web fluxes, food web stability, and the relative importance of ecological and stochastic processes shaping community assembly. Here, we quantified patterns and the relative distribution of organic matter fluxes through stream food webs spanning a natural ~25$^\circ$C temperature gradient. We then related these patterns to species- and community-level trait distributions of mean body size and population biomass turnover (*P:B*) within and across ecosystems. We predicted that 1) communities in warmer streams would skew towards smaller body size and higher *P:B* and 2) organic matter fluxes within warmer communities would increasingly skew towards smaller, higher *P:B* populations. Across the temperature gradient, warmer communities were characterized by smaller (~9% per $^\circ$C) body size and (~7% faster per $^\circ$C) *P:B* populations on average. Additionally, organic matter fluxes *within* warmer streams were increasingly skewed towards higher *P:B* populations, demonstrating that warming can restructure organic matter fluxes in both an absolute and relative sense. With warming, the relative distribution of organic matter fluxes also appeared to be increasingly 'non-random', suggesting stronger selection for traits driving high turnover with increasing temperature. Our study suggests that a warming world will favor energy fluxes through 'smaller and faster' populations, and that these changes may be more predictable than previously thought.

@@

<!--- warming and effects in food webs --->
<!--- Speeding up of food web: Species levels--->
Increasing global temperatures influence the provision and maintenance of ecosystem services by modifying the network of species interactions that underpin ecosystem functions [@brose2012; @woodward2010; @deruiter1995; @thompson2012]. Warming effects permeate across multiple levels of biological organization, from control on individual metabolic rates [@gillooly2001; @brown2004] and biological activity [e.g., attack rate, handling time, growth rates, etc.\; @dell2014], to broad-scale shifts in community assembly and structure [@saito2021; @gibert2019a; @nelson2017a]. Across global climate gradients, such temperature-induced changes have the potential to mediate food web stability [@baiser2019] by altering the acquisition and allocation of resources [@zhang2017] and the magnitude and relative distribution of energy fluxes in food webs [@may1972; @mccann1998; @barnes2018].

Understanding how warming alters energy flux in food webs requires information about how temperature modifies relationships between ecosystem structure and function. Such connections likely depend upon how warming influences functional trait distributions and how functional traits translate to patterns of energy demand [@loreau2001; @norberg2001; @norberg2004]. Additionally, warming can modify community assembly processes [@saito2021]---both deterministic [e.g., environmental/niche filtering, @whittaker1962] and stochastic [e.g., neutral theory, @hubbell2001]---leading to many possible relationships between warming, the relative abundance of species, and dominance of particular functional traits. For example, although temperature appears to have minimal control on species richness across taxonomic groups [e.g., @bastazini2021], warming can have profound effects on the dominance structure (i.e., evenness) of ecological communities, often by reducing community evenness and favoring a few well-adapted species [@hillebrand2008]. This strong environmental filtering is likely to skew trait distributions in natural communities [@therriault1999]. However, the relative distribution of traits can be further modified by additional processes unrelated to environmental filtering that alter species relative abundances [e.g., species interactions, @therriault1999;\; demographic stochasticity, @hubbell2001]. These additional processes can exaggerate or counter any skew in trait distributions driven by environmental filtering and thereby modify the relationship between species' traits and the absolute and relative energy demands in food webs.

<!--- Connect traits to the 'speed' of ecosystem energy fluxes--->
Body size is a fundamental trait that is influenced by temperature and has great potential to influence energy flux [@atkinson1994; @gardner2011; @daufresne2009]. Reduced body size in response to warming can arise at multiple levels of organization, including increased relative abundance of smaller species in warmed communities [@bergmann1848] or smaller individuals in warmer populations [@james1970], or reduced absolute body size of warmed individuals [@atkinson1994]. Thus, warmer communities are likely to contain both smaller species and individuals. These relationships can have important implications for energy and material flux in ecosystems because body size is a strong determinant of species life-history patterns [@peters1983; @zeuss2017; @altermatt2010; @nelson2020b], as well as developmental [@angilletta2004] and metabolic rates [@brown2004; @gillooly2001].

<!--- Write the main goals of the paragraph--->
Metabolic rate, in addition to being controlled by body size, is another key trait that is influenced by temperature and tied to variation in energy flux through food webs [@gillooly2001; @barnes2018]. Warming can influence metabolic rates indirectly through reductions in body size, as well as directly through its effects on subcellular kinetics [@bideault2019; @osmond2017]. These processes can modify ecosystem patterns through changes in population carrying capacity [@bernhardt2018], species' relative abundances and consequent effects on consumer-resource interactions or food web structure [@gibert2019a; @bideault2019]. Moreover, metabolic rates are also tied to many biological processes [@dell2014], including growth rate [@gillooly2001], developmental rate [@zuo2012; @nelson2020b], voltinism [@zeuss2017], and biomass turnover rate [@huryn2007; @brown2004]. Taken together, these changes suggest that warming should lead to a smaller, faster world in which ecosystem processes accelerate through the effects of smaller body size and higher turnover rates, with a potentially strong imprint on food web structure and energy fluxes [@gibert2019a].

We quantified patterns of organic matter flux in stream food webs across a natural temperature gradient (\~5--28$^\circ$C) in southwestern Iceland. Previous research in these streams has shown a strong positive effect of warming on primary production both among streams [@demars2011; @padfield2017] and seasonally within streams [@ogorman2012; @hood2018]. Invertebrates in these streams rely on autotrophs [@ogorman2012; @nelson2020], thus the dynamics of primary production functions as a strong control on energy flow through consumers [@junker2020]. We therefore predicted that annual organic matter fluxes to consumers would increase with stream temperature, mirroring patterns in resource availability and consumer energy demand. We also hypothesized that temperature would act as a principle environmental filter on community assembly and organic matter fluxes by favoring 'fast' life-history traits associated with small-bodied organisms. Specifically, we predicted that warming temperatures would lead to reduced average body size and increased average biomass turnover (i.e., *P:B* ratio) of populations among streams. We also predicted that *within* communities, organic matter fluxes would be dominated by small-bodied and high *P:B* taxa at higher temperatures, and that these patterns would not arise by random change [i.e., a 'selection effect' *sensu* @huston1997]. Our results should help refine general predictions about how ongoing climate warming, and its influence on key traits, is likely to shape energy flux through animal food webs.

@@

We studied six streams within the Hengill geothermal field of southwestern Iceland (64$^\circ$ 03'N 021$^\circ$ 18'W) that varied in mean annual temperature from ~5 to 30$^\circ$C. Hengill is
characterized by indirect geothermal heating of groundwater [@arnason1969], leading to natural variability in stream temperatures (4.5--54.0 $^\circ$C), but similar solute
chemistries [@friberg2009]. These conditions create a "natural laboratory" for isolating the effects of temperature on ecosystem processes [@ogorman2014; @nelson2017a]. We selected streams to maximize the temperature range, while minimizing differences in the structural aspects of the primary producer community [@junker2020a]. In each stream, we measured temperature and water depth every 15 min from July 2010 through August 2012 (U20-001-01 water-level logger, Onset Computer Corp., Pocasset, MA, USA). Light availability in the watershed was measured every 15 minutes from atmospheric stations (HOBO pendant temperature/light UA-002-64, Onset Computer Corp., Pocasset, MA, USA).

@@

Daily secondary production of invertebrate taxa was calculated using the instantaneous growth rate method [IGR, @benke2017]. Growth rates were determined using taxon-appropriate
approaches described in Junker and others [-@junker2020]. Briefly, growth rates of common taxa (e.g., Chironomidae spp., *Radix balthica*, etc.) were measured using *in situ* chambers [@huryn1986].
Multiple individuals (*n* = 5--15) within small size categories (\~1-mm length range) were photographed next to a field micrometer, placed in the stream within pre-conditioned
chambers for 7--15 days, and removed and photographed. Individual lengths were measured from field pictures using image analysis software [@schindelin2012], and body
lengths were converted to mass (mg ash-free dry mass [AFDM]) using published length-mass regressions [@benke1999; @ogorman2012; @hannesdottir2013]. Growth rates (*g*, d^-1^) were
calculated by the changes in mean body size (*M*) over a given time interval (*t*) with the following equation:

@@

Organic matter fluxes (mg AFDM m^-2^ y^-1^) through the community were calculated using the trophic basis of production method [TBP\; @benke1980]. Taxon-specific secondary production estimates were combined with diet proportions (see Supplemental Materials), diet-specific assimilation efficiencies, *AE~i~*, and net production efficiencies, *NPE* [e.g., @mccullough1975], to estimate consumption of organic matter. For each food
category, *i*, diet proportions were multiplied by the gross growth efficiency ($GGE_{i} = AE_{i} * NPE$) to calculate the relative production attributable to each food
category. The relative production from each food type was then multiplied by the interval-level production and finally divided by $GGE_{i}$ to estimate consumption of organic matter from each food category by each taxon [@benke1980]. Consumption was calculated for each taxon across sampling intervals (typically \~1 month). Total interval consumption was calculated by summing across all taxa, while annual consumption was calculated by summing across all taxa and intervals. Variability in consumption estimates was estimated through a Monte Carlo approach, wherein bootstrapped vectors of secondary production for each taxon (see *Secondary production* methods above) were resampled and consumption estimated with the TBP method using modeled diet proportions (see *Diet analysis* in Supplemental Materials), diet-specific assimilation efficiencies, and net production efficiency. Variability in *AE~i~* was incorporated by resampling values from beta distributions fit to median and 2.5% and 97.5% percentiles for each diet item: diatoms = 0.30 (95% percentile interval (PI): 0.24-0.36), filamentous and green algae = 0.30 (95% PI: 0.24-0.36), cyanobacteria = 0.10 (95% PI: 0.08-0.12), amorphous detritus = 0.10 (95% PI: 0.08-0.12), vascular and non-vascular plants (bryophytes) = 0.1 (95% PI: 0.08-0.12), and animal material = 0.7 (95% PI: 0.56-0.84)[@welch1968; @cross2007; @cross2011; @benke1980; @benke1997]. Variability in *NPE* was incorporated by resampling values from an assumed beta distribution with median *NPE* = 0.45 (95% PI = 0.4-0.5). Beta distributions were fit in R [@rcoreteam2022] using the 'get.beta.par()' function within the *rriskDistributions* package [@belgorodski2017].

@@

We predicted that warming would favor taxa with smaller body size and higher population turnover and therefore OM fluxes would be skewed towards small body size (*M*) and higher *P:B* across and within communities. To assess the potential importance of environmental filtering across communities, we used bootstrapped linear regressions to quantify the relationship between mean annual temperature ($^\circ$C) and either mean population body size ($\overline{M}$) or biomass turnover ($\overline{P:B}$) of each stream community. Here, `r format(temperature_stats[['n_boot']], big.mark = ",", scientific = FALSE)` values of population *M* or *P:B* were resampled with replacement from population secondary production vectors (see *Secondary Production* above) for each taxon within each stream. The mean of all populations within a stream was calculated and a linear model was fit between *log~e~*-transformed $\overline{M}$ or $\overline{P:B}$ and mean annual temperature. Response variables were transformed to meet the assumption of normally distributed residual variation.

@@

To examine whether and how temperature influences the skew of OM flux among populations within each community, we quantified the probability of observing the skewed distributions in OM fluxes by random chance. The feasible range of skewness values within a community is inherently tied to the evenness of OM fluxes. This dependence on evenness can make it difficult to determine the importance of random versus ecological processes on the distribution of OM fluxes by comparison of raw skewness measures alone. We predicted that species $M$ and $P:B$ would be increasingly important traits structuring OM fluxes within a community and therefore warmer streams would exhibit highly skewed OM distributions that would be unlikely due to chance (i.e., 'non-random ordering'). In contrast, cooler streams would exhibit OM fluxes with skew values that are more likely due to random chance ('random ordering'), regardless of raw skewness values, suggesting other traits or processes govern the distribution of OM fluxes within their communities. To accomplish this, we first had to account for statistical constraints that restrict the range of possible outcomes [i.e., feasible set, @haegeman2008; @diaz2021a], given the number of species and the relative distribution of OM fluxes within a community. The number of unique orderings of species increases to computationally intractable numbers very quickly (e.g., $S!$ for 10 species $\approx 3.6 \times 10^{6}$ unique orderings). Therefore, we chose to permute a random subset of each stream community's feasible set by randomly ordering species and calculating skewness in the cumulative distribution of annual OM fluxes 100,000 times in each stream. This permuted set allowed us to calculate the probability of observing the empirical skewness value, $Sk_{flux}$, compared to a random ordering given the distribution of relative OM flux. We assessed the likelihood of non-random ordering as the distance from the central mass of the random skew distributions within each stream. Therefore, communities in which OM fluxes are likely organized non-randomly are indicated by observed skew values near the tails of these random distributions.

@@

We compared our skew estimates (i.e., $Sk_{flux}$) of OM flux based on ordered data (i.e., ordered from small to large *M* or low to high *P:B* ratios; Figure 3B and Figure 4B) to skew estimates based on randomly sorted data to detect whether OM fluxes through smaller and higher *P:B* taxa could be attributed to these traits vs. random community assembly processes. The probability of observing a similar or more extreme skew of OM fluxes in relation to body size was variable among streams and ranged from `r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_mean")) %>% round(2)` (`r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_quant2.5")) %>% round(2)`--`r m_skew_summ %>% slice_min(M_skew_prob_mean) %>% select(matches("M.*prob_quant97.5")) %>% round(2)`; 95% PI) to `r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_mean")) %>% round(2)` (`r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_quant2.5")) %>% round(2)`--`r m_skew_summ %>% slice_max(M_skew_prob_mean) %>% select(matches("M.*prob_quant97.5")) %>% round(2)`); there was a very weak positive association between this probability and temperature (Figure 3C). In contrast, the probability of a similar or more extreme skew in relation to *P:B* ratios ranged from `r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_mean")) %>% round(2)` (`r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant2.5")) %>% round(2)`--`r pb_skew_summ %>% slice_min(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant97.5")) %>% round(2)`) to `r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_mean")) %>% round(2)` (`r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant2.5")) %>% round(2)`--`r pb_skew_summ %>% slice_max(pb_skew_prob_mean) %>% select(matches("pb.*prob_quant97.5")) %>% round(2)`; Figure 4C) and in this case there was a clear trend towards a more structured OM flux distribution---i.e., favoring fluxes through high *P:B* taxa--at warmer temperatures. The likelihood of elevated fluxes among high *P:B* taxa became much more likely with warming, and the probability of random ordering decreased by `r round(pb_prob_temp_median, 3)*100`% (`r round(pb_prob_temp_quant2.5,3)*100`%--`r round(pb_prob_temp_quant97.5,3)*100`%) for every 1$^\circ$C increase in temperature (Figure 4C).

@@

# Conflict of Interest

The authors have no conflict of interest to declare.

@@

Figure 1

@@

Figure 2
